\chapter*{Abstract}
JavaScript provides very few operations to process its arrays.
Additionally, these operations are inconsistent: some modify the existing
array, but others create a new one. The programming language Haskell provides a
much more extensive and concise collection of such operations for
lists~\cite{haskell_list}. Additionally, its powerful type system can express
constructs (called type classes) that enable the definition of abstract
operations usable with many different types.\\ 
This project, therefore, investigated how to port various functionalities of
the Haskell standard library for lists to JavaScript. Furthermore, the focus
was to figure out how to model type classes analogous to Haskell using "JSDoc",
the optional type system of JavaScript~\cite{jsdoc_use_2023}. The solutions and
findings of this project will become part of the existing zero-dependency Web
UI Toolkit "Kolibri"~\cite{kolibri}. \\
Since Haskell is a functional programming language, it uses many concepts that
are not very familiar in JavaScript. However, JavaScript supports some of them,
which, therefore, have been used where possible. Using them allows the writing
of robust, well-testable, and reusable code. Unit and user tests ensured the
correctness and usability of the artifacts. \\
JavaScript defines iteration protocols to process data structures element by
element. This project exploited these protocols to define a new data structure
called "sequence", and operations to transform and process it. These operations
are compatible with any data structure that complies with the iteration
protocols. All functions evaluate their input lazily and do not mutate it.\\
Using JSDoc, the type class monad could be defined. Limitations lie, however,
in the missing name overloading in JavaScript and the missing higher-kinded
types in JSDoc.\\
Based on these findings, JINQ emerged, which allows any data structures
implementing the type class monad to be queried uniformly.\\
In his paper "Why functional programming matters," John Hughes shows how
functional programming enables developers to write reusable and modular
code~\cite{hughes_why_1989}. He explains this with several example programs.
The results of this project allow the implementation of the programs showcased
in Hughes' paper! Additionally, JINQ shows that JSDoc allows to define
monads in JavaScript. Thus, this project reveals that JavaScript can
effectively implement and utilize many concepts known from functional
programming.
